Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-09-11 Response_to_Reviewers.tex

% Thank you for submitting your manuscript to Computers, Environment and Urban Systems. I have completed the review of your manuscript and a summary is appended below. The reviewers recommend reconsideration of your paper following major revision. I invite you to resubmit your manuscript after addressing all reviewer comments.

% When revising your paper, pay special attention to the explicit presentation of every model that is investigated in the paper. Present each model as an algorithm/unambiguous pseudocode and supply the Netlogo code for downloading (all this can be in appendix). The Schelling model, for example, has several Netlogo implementation, and it is critically important to know which of them is investigated. The same is true for the Sugarscape model. See comments of the reviewer #1 in this respect.

% When resubmitting your manuscript, please carefully consider all issues mentioned in the reviewers' comments, outline every change made point by point, and provide suitable rebuttals for any comments not addressed.

As a response to several reviewers, we made the following general changes:

\begin{itemize}
	\item Several comments, from the first and third referees especially, suggested that we clarity and explicit the description of our two models: we clarified the descriptions and justified our choices more thoroughly, and also provided pseudo-codes for each model in the Appendix so as to help the reader navigate the implementation chosen compared to other articles. Finally, we mentioned more clearly in the text the fact that source codes are openly available in a github repository.
    \item There was a conflict between the comments of the third referee and the other referees about the comparison of phase diagram. In order to conciliate as much as possible the different parties, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper.
    \end{itemize}

@@

\medskip

We now turn to point-by-point responses to referees comments.

@@

\medskip

Concerning specifically the comments made by the first referee:

\begin{itemize}
	% 1. I think it would be better if the author would focus on a single model. This way they could describe the model and the analyses in more detail. As it stands, the models are treated as black boxes. 
	\item The suggestion to focus on a single model is a qualified comment. However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies rather than with a single model. In that perspective, the ``black-box'' is seen as an advantage as it reinforces genericity. We acknowledge the fact that this choice led us to shorten the model presentations. Nevertheless, we have taken notice of your remark by improving the description of the models and their results through the addition of pseudo-codes in Appendix, result descriptions in the text and by recalling more clearly the link to the model source we used. Furthermore, we have focused on some of the most classical models of the ABM literature so as to lower the reader's cost of entry into the methodology.
    %We believe this is sufficient as the models have not been adjusted from these textbook cases implementations.

   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
   \item \comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..} \comment[CC]{peut-être en prenant des cas extrêmes sur Sugarscape?}

	% 2. It is very important to describe the models. This is especially relevant for the Schelling model, which has many variants; each has different dependencies on the initial pattern. In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
	\item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project. This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.% It does not reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
	 \comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants? Comment ça? Bien sûr qu'il le fait (sous contrainte du taux de vacance).
	  Cela dit y'a bien un point sur lequel l'implem Netlogo n'est pas bonne: c'est que la relocalisation des mécontents est random alors que dans le bouquin de Schelling, apparemment, c'était dans un voisinage proche.}

    \item More generally concerning the choice of these two models, we aim at helping the readers to distance themselves from the case studies and focus on the genericity of the approach. This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs. This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.

@@

	% 3. I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.
	\item Furthermore concerning the choice of the Sugarscape model, we think it is a good example of a stylized model of a spatialized socio-economic system, and according to (Batty, 2005) a good metaphor of an urban system. Our choice was accordingly clarified in the paper.
%: "In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants." % pas besoin de mettre les modifs ici

	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
	%In order to clarify the concept of ``meta-parameter'' that we prefer to keep, we now propose a short definition after the first use of the term (section Methods): "In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process".
    \item \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
    \comment[CC]{d'acc.}

	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
	\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}

@@

	% 6. Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns
	\item \comment[JR]{biblio a rajouter}

	% 7. Line 238: Please explain why this particular index was chosen and discuss its properties. Connect the discussion to other indices like L1, and Linf norms, etc. Would we get qualitatively different results from different indices? 
	\item \comment[JR]{on avait teste avec le Monge ca donnait a peut pres pareil ; donc les normes alpha surement encore plus. je regarde.}

	% 8. Regarding the spatial generator: the authors should explain why this generator is relevant to these two models? 
	\item This spatial generator is relevant to both model as it produces density grids at a ``metropolitan scale'' (see detailed discussion added in Appendix and the original paper from which we used the generator), which is the scale at which both model were initially intended to be. We made this aspect more explicit.

	In the case of Schelling, this is the scale at which most empirical segregation indices are computed. In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty....), or to a generic landscape where a resource is grown otherwise. In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density, and that not acknowledging this diversity leads to potentially false results. Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most harmful one (e.g. the least realistic, although this is the case of most Schelling models).
	\comment[CC]{argumentaire à ajouter dans le papier}

@@

	% 9. The authors should illustrate the contribution of each parameter of the generator to the generated initial pattern using a figure. Details of the generators should be included in an appendix
	\item The behavior of the spatial generator was described with more details in Appendix.

	% 10. Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? 
	\item Yes, we allow more agents to be in the same cell. The potential density is defined by the density grid generated, and is population with the same number of agents in all grids. This distribution is the one at initialisation. If the potential density is not reached, more agents can move in the cell, otherwise it is deemed full and unavailable for further moves. The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.
	\comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
	\comment[CC]{ça va comme réponse? }
	\comment[CC]{argumentaire à ajouter dans le papier}

	% 11. Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study. 
	\item cf infra. 
	\item density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution. Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling. We chose to use actual distribution of European cities to constraint our density generation.
\comment[CC]{argumentaire à ajouter dans le papier}

	% 12. Line 641: NetLogo supports grids, continuous space, and GIS elements. It does not restrict the user in any way.
    % \comment[JR]{plutot d'accord avec lui mais parce que proficient en NetLogo, peut etre que des debutants sont plus contraints ? quelquun aurait une ref là dessus ?}
    %\comment[MLT]{En fait, c'est surtout le cas des pre-built simulations proposées dans la librairie de NetLogo. J'ai un peu cherché, je trouve pas de références qui montrent explicitement la surabondance des grilles dans les modèles hors librairie...}[(JR) ok yes en effet ! cela devrait suffire comme argument ?]
	\item We agree that NetLogo does not restrict the use on what implementation to use. However, some features are much more easier as raster or the contrary, which could influence the choice of beginners. Furthermore, models in the built-in NetLogo model library mostly rely on rasters. We clarified that statement in the paper.

@@

Concerning the comments made by the second referee:

%This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.

%I also have some concerns:

\begin{itemize}
	% In the abstract the authors claim that initial spatial conditions are often taken for granted. I do not agree to this statement. Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.
	\item We forcefully agree with the referee in that initial spatial conditions can have a large impact on the results of the model. Maybe not all papers take them for granted, but what we meant is that compared to other parameters, initial spatial conditions are much less often explored using sensitivity analysis. Actually, we know only a few papers which have explored the impact of initial spatial conditions as a systematic criterion in a sensitivity analysis of model results. We clarified this point in the introduction.
    \comment[JR]{il s'agirait de nuancer ici}
\comment[CC]{argumentaire à ajouter dans le papier}

@@

	% Page 2 line 88: the authors state that “Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density”, I do not think this is true. Many models generate a heterogeneous synthetic population that is based on local data (replicating the actual population as closely as possible) and in far more detail then what is done in this paper. I am thinking of work in the line of disease diffusion, crime etc.
	\item \comment[JR]{il a un point car ce sont des classes de modeles qu'on a pas trop cité ? par contre la sensibilite a ces distribs precises n'est pas testée, ou pas systeématiquement : il faut qu'on clarifie ou reformule}

	% Page 2 line 98: “However, this step is rarely explicit and authors tend to generalize the conclusion of the model from a single, spatial case study”. I do not understand the first part of the this statement. The steps taken are often very explicit. Due to the amount of time it takes to correctly represent all spatial data, including population densities and the variation of other population variables over space, many authors indeed restrict themselves to a single case study due to the time intensity of this task. Many agent-based models are also not valid for other areas.
	\item \comment[JR]{idem la on parle pas exactement du meme type de modeles, mais son exemple prouve notre propos : le modele tourne sur une situation unique, et on teste pas l'effet de l'espace.}

	% The authors do not differentiate between data driven empirical models versus theory-driven models. I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.
	\item \comment[JR]{oui du coup en disant ça ça repond à ses deux points précédents}

	% In section 1.3 (Previous considerations on the effects of the spatial configuration in simulation models) authors provide a lot of evidence on previous work on the impact of initial spatial conditions on the output of models. This is contrary to the statement in the abstract that initial conditions are often taken for granted.
	\item Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled. But first, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition. Second, the references cited mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but none provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper. We have tried to argue this point at the end of the section.
    \comment[CC]{argumentaire à ajouter dans le papier}

### 2018-09-11 Response_to_Reviewers.tex

	% The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs. As these are vectors, they are not so different from time series and many techniques exist to compare vectors in this domain. I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.
	% -> 
	\item

	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
	\item The functionalities of OpenMOLE we used were detailed; we did not use genetic algorithms but sampling methods and high performance computing environments.
    \comment[CC]{argumentaire à ajouter dans le papier}

	% In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions. Actually this came as a surprise to me. I had the impression that all phase diagrams would be evaluated against all other diagrams. Perhaps this should be explained earlier in the paper.
	\item

	% Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.

	\item \comment[JR]{Clem tu es d'accord là dessus ?}
    \comment[CC]{Non mais il chipote complètement à ce niveau! Vu la figure pour moi c'est totalement équivalent dans un sens et dans l'autre, le tout c'était qu'il y'a une double descente chromatique et y'a même pas de densité = rouge ou quoi puisque c'est un rapport. Si tu veux lui faire plaisir on change mais je vois pas trop la nécessité.}

	% I appreciate the fact that the authors in section 4 clearly indicate the limitations of their work (and I share the limitations listed under “comparing phase diagrams”). Personally I do not think that releasing the density grids is an important opportunity. As mentioned before I think that the core of the work is in the comparison of the phase vectors. I am not so sure the second opportunity listed in 4.2 is very valid. I do not think that the CEUS journal targets social sciences exactly.
	Thanks for the remark. We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
	\comment[CC]{argumentaire à ajouter dans le papier}

### 2018-09-11 Response_to_Reviewers.tex

\medskip

Concerning specifically the comments made by the first referee:

\begin{itemize}
	% 1. I think it would be better if the author would focus on a single model. This way they could describe the model and the analyses in more detail. As it stands, the models are treated as black boxes. 
	\item The suggestion to focus on a single model is a qualified comment. However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies rather than with a single model. In that perspective, the ``black-box'' is seen as an advantage as it reinforces genericity. We acknowledge the fact that this choice led us to shorten the model presentations. Nevertheless, we have taken notice of your remark by improving the description of the models and their results through the addition of pseudo-codes in Appendix, result descriptions in the text and by recalling more clearly the link to the model source we used. Furthermore, we have focused on some of the most classical models of the ABM literature so as to lower the reader's cost of entry into the methodology.
    %We believe this is sufficient as the models have not been adjusted from these textbook cases implementations.

   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
   \item \comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..} \comment[CC]{peut-être en prenant des cas extrêmes sur Sugarscape?}[(JR) je regarde...]

	% 2. It is very important to describe the models. This is especially relevant for the Schelling model, which has many variants; each has different dependencies on the initial pattern. In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
	\item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project. This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.% It does not reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
	 \comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants? Comment ça? Bien sûr qu'il le fait (sous contrainte du taux de vacance).
	  Cela dit y'a bien un point sur lequel l'implem Netlogo n'est pas bonne: c'est que la relocalisation des mécontents est random alors que dans le bouquin de Schelling, apparemment, c'était dans un voisinage proche.}[(JR) ah oui cest selon la vacance ; mais il y a des regimes ou ca converge jamais sous forme de cluster et ca reste un etat dynamique sans structure ; oui par contre on a du reprendre la random reloc, c'est ce que font les physiciens aussi]

@@

	% 3. I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.
	\item Furthermore concerning the choice of the Sugarscape model, we think it is a good example of a stylized model of a spatialized socio-economic system, and according to (Batty, 2005) a good metaphor of an urban system. Our choice was accordingly clarified in the paper.
%: "In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants." % pas besoin de mettre les modifs ici

	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
	%In order to clarify the concept of ``meta-parameter'' that we prefer to keep, we now propose a short definition after the first use of the term (section Methods): "In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch). We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process".
    \item \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
    \comment[CC]{d'acc.}

	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
	\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}[(JR) il me semble que les distribs sont assez centrées et que ca revient quasiment au même ; par contre Clem si tu as pris median pour les stats et moi mean pour les plots c'est pas fou ; tu as encore le code des analyses stats ? (c'est sur le git ?)]

@@

%This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.

%I also have some concerns:

\begin{itemize}
	% In the abstract the authors claim that initial spatial conditions are often taken for granted. I do not agree to this statement. Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.
	\item We forcefully agree with the referee in that initial spatial conditions can have a large impact on the results of the model. Maybe not all papers take them for granted, but what we meant is that compared to other parameters, initial spatial conditions are much less often explored using sensitivity analysis. Actually, we know only a few papers which have explored the impact of initial spatial conditions as a systematic criterion in a sensitivity analysis of model results. We clarified this point in the introduction.
    \comment[JR]{il s'agirait de nuancer ici}
\comment[CC]{argumentaire à ajouter dans le papier}

	% Page 2 line 88: the authors state that “Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density”, I do not think this is true. Many models generate a heterogeneous synthetic population that is based on local data (replicating the actual population as closely as possible) and in far more detail then what is done in this paper. I am thinking of work in the line of disease diffusion, crime etc.
	\item \comment[JR]{il a un point car ce sont des classes de modèles qu'on a pas trop cité ? par contre la sensibilite a ces distribs precises n'est pas testée, ou pas systeématiquement : il faut qu'on clarifie ou reformule}

### 2018-09-18 paper.tex
Update on Overleaf.

\usepackage{xparse}
\usepackage{ifthen}
\DeclareDocumentCommand{\comment}{o m o o o o}
{\ifthenelse{\draft=1}{
  \IfValueT{#1}{
      \textcolor{red}{\textbf{C (#1) : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{ForestGreen}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{Aquamarine}{\textbf{A4 : }#6}}
    }
    \IfNoValueT{#1}{
      \textcolor{red}{\textbf{C : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{ForestGreen}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{Aquamarine}{\textbf{A4 : }#6}}
    }
 }{}
}

@@

Computer simulation has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}\added{, what has been synthesized as the emerging field of geosimulation by \cite{benenson2004geosimulation}.} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems. The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model in understanding the role of space on social interactions and environmental processes, in a city for example. We think simulation can become a very good tool to achieve this, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is represented. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\

@@

Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in the most sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic densities. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of the urban processes under inquiry because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers represent space is therefore a central element of geographical simulation models. However, this step is rarely explicit and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one. A meaningful way of representing space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour in plausible spatial arrangements.

@@

 Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
%\cite{ligmann2013spatially}

@@

\todo{explain why this generator is relevant for both models (discussion on scale ?)}

@@

@@

In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised correlates to a certain extent with the level of segregation measured at the end of the simulation run. Indeed, compared to compact density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents. Discontinuous grids have the same effect, although attenuated. The results obtained with Moran's I are opposite, because this index measures spatial autocorrelation at the global level and that compact cities have higher levels of global autocorrelation by construction. However, linear models with and without the type of density distribution yield the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant. The similarity of the coefficient in both cases means that the effect of the model's parameters (and thus the mechanism by which agents of similar group cluster in space) is the same regardless of the distribution of density. The way polycentric and discontinuous density grid exhibit higher segregation is by allowing buffer zones of low density to surround pockets of homogeneity, which is impossible in a compact city, because everyone is at reach of everyone else. The buffering process confirms previous results obtained with network structures \citep{Banos2012} and supports the conclusion that space acts here on top of mechanisms rather than in interaction with them.

@@

As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions, that become senseless taken out of context. Geography and the social sciences in general have been in the last decades strongly influenced by insights from physics, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space. We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.

>>>>

\usepackage{xparse}
\usepackage{ifthen}
\DeclareDocumentCommand{\comment}{o m o o o o}
{\ifthenelse{\draft=1}{
  \IfValueT{#1}{
      \textcolor{red}{\textbf{C (#1) : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
    }
    \IfNoValueT{#1}{
      \textcolor{red}{\textbf{C : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
    }
 }{}
}

@@

Computer simulation has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}\added{, what has been synthesized as the emerging field of geosimulation by \cite{benenson2004geosimulation}.} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems. The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model in understanding the role of space on social interactions and environmental processes, in a city for example. We think \replaced{that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions}{simulation can become a very good tool to achieve this}, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is represented. Unfortunately, the first condition is not always met, and the second is seldom even mentioned. This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions. To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\

@@

Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in the most sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic densities. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of the urban processes under inquiry because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers represent space is therefore a central element of geographical simulation models. However, this step is rarely explicit. For example

and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one. A meaningful way of representing space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour in plausible spatial arrangements.

@@

 Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) \replaced{or}{either} on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time, \added{offering a systematic inquiry into the impact of systematically defined spatial inputs on the model outputs. More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.}

%\cite{ligmann2013spatially}

@@

\added{The spatial generator and its resulting grids are relevant to the case study models we have picked (Schelling and Sugarscape) because it produces density grids at a ``metropolitan scale'', which is the scale at which both model were initially intended to be. In the case of Schelling's segregation model for example, this scale is the one at which most empirical segregation indices are computed and compared to the model outputs. In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources [REF BATTY], or to a generic landscape where a resource is grown otherwise. In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density and that acknowledging this diversity might leads to a variation in the model outputs. Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most unrealistic one, although unfortunately the most common one in Schelling implementations.}
\todo{add ref to Batty's paper}

@@

\added{In general, the implementations of Schelling models allow only one agent per cell, and their initial distribution is random, therefore following a uniform distribution across the modelled city. In this experiment, we allow more than one agent to be in a given same cell. The potential density of a cell is defined by the density grid generated. If the potential density of a cell is not reached at initialisation, more agents can move into the cell during the course of the simulation, otherwise it is deemed full and unavailable for movers. The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells. Empirical distributions of density in cities are important in our framework because we want to test models with realistic ranges of initial patterns of density distribution. Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling. We chose to use actual distribution of European cities to constraint our density generation.}

@@

In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised correlates to a certain extent with the level of segregation measured at the end of the simulation run. Indeed, compared to \added{the reference case of} compact \added{(monocentric)} density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents. Discontinuous grids have the same effect, although attenuated. The results obtained with Moran's I are opposite, because this index measures spatial autocorrelation at the global level and that compact cities have higher levels of global autocorrelation by construction. However, linear models with and without the type of density distribution yield the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant. The similarity of the coefficient in both cases means that the effect of the model's parameters (and thus the mechanism by which agents of similar group cluster in space) is the same regardless of the distribution of density. The way polycentric and discontinuous density grid exhibit higher segregation is by allowing buffer zones of low density to surround pockets of homogeneity, which is impossible in a compact city, because everyone is at reach of everyone else. The buffering process confirms previous results obtained with network structures \citep{Banos2012} and supports the conclusion that space acts here on top of mechanisms rather than in interaction with them.

@@

As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions. Geography and the social sciences in general have been strongly influenced by physics in the last decades, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space. We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.
